Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-22 latex/main.tex
354 vs 356
Closes #700

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 356 have at least four loci supporting the non-focal parent on both sides of the breakpoint
(we refer to these as ``robust'' recombination events,
Figure~\ref{fig:recombinant_qc}; see STAR Methods)
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass the robustness filtering criteria (Figure~\ref{fig:recombinant_qc}).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

@@

In addition to the analysis of the 356 ``robust'' recombination events with a minimum of four nucleotides supporting each parent,
we analyzed the relationship between the full set of 855 recombinants and case numbers.
These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022).
Again, the number of recombinant events rose with the global number of cases
(linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
and with the diversity of lineages in each week
(linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
The single best predictor of recombination events among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
was again the product of cases and diversity ($n \; H$),
with an adjusted $R^2$ of 26.9\%,
although this was now followed closely by expected heterozygosity on its own ($H$),
with an adjusted $R^2$ of 26.5\%.
Thus, similar conclusions are reached with either dataset,
although filtering the recombination events to the 356 with better support revealed a stronger relationship with case numbers.

>>>>

Using sc2ts, we built an ARG of $\sim$2.48 million samples spanning the COVID-19 pandemic.
Inference took 21 days on a server equipped with dual EPYC (details) and 500 GB of RAM 
(~5 CPU years; Figure~\ref{fig:inference_resources}).
The ARG contains 2,689,054 nodes representing the genealogical relationships
among 2,482,157 sampled genomes in the Viridian v04 dataset,
which were collected from January 01, 2020 to February 20, 2023 \cite{Hunt2024}.
Of the nodes, 855 represent recombination events,
of which 354 pass quality control (QC) filters
(see STAR Methods, Figure~\ref{fig:recombinant_qc}),
reflecting the overwhelmingly tree-like structure of SARS-CoV-2 evolutionary history.

@@

We find strong evidence for the recombination events leading to 17 Pango X lineages
in the ARG (Table~\ref{tab:pango_x_lineages}),
all of which pass QC (STAR Methods).
%% JK: Does saying this here help at all? I think we may be better off without 
% These events are supported by high numbers of averted
% mutations (Table 1) and high numbers of loci matching their suggested parents
% (see their copying patterns in Figure SX). 
Among these Pango X lineages are XA,
which is the first designated Pango X lineage that
showed signs of onward transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation
(Figure~\ref{fig:subgraphs}A; Table~\ref{tab:pango_x_lineages}), 
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

@@

In addition to the analysis of the 354 recombination events 
that pass QC,
we analyzed the relationship between the full set of 855 recombinants and case numbers.
These additional recombinants occurred disproportionately during the Delta and first Omicron wave (July, 2021 to February, 2022).
Again, the number of recombinant events rose with the global number of cases
(linear regression: $3.1 + 5.3 \; 10^{-7} x$, $p = 0.0006$)
and with the diversity of lineages in each week
(linear regression: $0.42 + 18.3 x$, $p = 2 \; 10^{-12}$).
The single best predictor of recombination events among the predictors tested ($n$, $H$, $n \; H$, $n^2$, $n^2 H$)
was again the product of cases and diversity ($n \; H$),
with an adjusted $R^2$ of 26.9\%,
although this was now followed closely by expected heterozygosity on its own ($H$),
with an adjusted $R^2$ of 26.5\%.
Thus, similar conclusions are reached with either dataset,
although filtering the recombination events to the 354 with better support revealed a stronger relationship with case numbers.
